The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}}, say \$20.00, and eventually
produces another \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted line
with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type {{a+0}}; we indicate this value
  as {{left 20}};
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
\roshan{Why was this deleted?}
The example illustrates that the empty type and negative types have a
computational interpretation related to continuations: negative types denote
values that backtrack to satisfy dependencies, or in other words act as debts
that are satisfied by the backward flow of information.

@@

As before, the execution begins at line~(0) with the value {{20}}. At
line~(1) two values, {{20}} and {{()}}, flow forward. One can think of the
value {{()}} (of type {{1}}) as ``having a credit card.'' The credit card
isn't money, nor is it debt, but is the option to generate a credit-debt
constraint.  At line~(2) we exercise this option and hence have three values:
the initial value {{20}} flowing from line (1) and two entangled values,
{{1/alpha}} and {{alpha}}. The {{alpha}} and {{1/alpha}} are unspecified
values, i.e., we don't yet know how much money we need to borrow, but we do
know that what is borrowed must be what is returned. Hence~{{alpha}} denotes
the presence of an unknown quantity and dually {{1/alpha}} should be thought
of as the absence of an unknown quantity.
At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
fixing values of all three branches, i.e., we have paid our debt of the missing
information and the third subcomputation has the known value {{20}}. After
payment of our debt, as line (4) shows, we have use of our credit card again.

\roshan{Also, why was this deleted?}
Unlike with negative types, wherein only one value existed at a time and the
computation backtracked, here we have three values that \emph{exist at the
  same time}. In other words, the computation with fractions is realized with
a schedule in which every value independently and concurrently proceeds
through its subcomputation.

@@

\noindent 
\textit{Example.}  We use the type {{bool}} as a shorthand to denote
the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
to be {{false}}. The following combinator is represented by the given
diagram:

{{c : b * bool <-> b + b}}

@@

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
%% <<<<<<< local
%% categorical semantics informally via the graphical language. The
%% operational semantics is presented in Sec.~\ref{sec:rat}. As will be
%% detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
%% compact closed structure provides several properties of interest: (i)
%% morphisms or wires are allowed to run from right to left, (ii) the
%% structure admits a trace that, depending on the situation, can be used
%% to implement various notions of loops and
%% =======
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

As will be detailed in the remainder of this section, the compact closed
structure provides several properties of interest: (i) morphisms or wires are
allowed to run from right to left, (ii) the structure admits a trace that,
depending on the situation, can be used to implement various notions of loops
and
%% >>>>>>> other
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual
operator is an involution, and (iv) the structure is equipped with
objects representing higher-order functions and a morphism
\textit{eval} that applies these functional objects. Interestingly
each monoidal structure provides the same ingredients, resulting in
two dualities, two traces, two involutions, and two notions of
higher-order functions.

@@

%% <<<<<<< local
%% Hence, the intuitive way to read the diagrams is that they reverse the
%% flow of information -- i.e. they capture a notion of \emph{negative
%%   information flow}. In normal {{langRev}} circuits, information flows
%% from left-to-right. The operator {{eps+}} causes information to flow
%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
%% it turns a right-to-left flow of information in a circuit into a
%% left-to-right flow.  The diagram below is an example of using {{eta+}}
%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
%% from the introduction. (Here the {{0}} wires have been shown for
%% completeness.)
%% =======
\paragraph*{Negative Information Flow.} 
The intuitive way to read the diagrams is that they reverse the flow of
information. In previous {{langRev}} circuits, information was always flowing
from left to right. The presence of {{eps+}} operations cause information to
flow from right to left which can be reversed again using {{eta+}}.  As an
example, the diagram representing the isomorphism from the introduction is:
%% >>>>>>> other
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/algebra1.pdf}
}
\end{center}

% \paragraph*{Time Traveling.}
%% <<<<<<< local
%% =======
% A time traveling intuition is also applicable here: if view the
% left-to-right direction of information flow as the canonical forward
% direction of information, then the action of {{eps+}} allows values,
% aka information particles, to flow backwards in time. Particles that
% flow backwards in time get to see and interact with the history of
% particles that they coexist with, i.e., values that they exist
% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
% (whose construction is shown in Section \ref{sec:neg-constructions}).
%% >>>>>>> other

%\subsection{Compact Closure with Additive Inverses}
%% <<<<<<< local
%% \emph{Coherence for Compact Closure with Additive Duality.}  The
%% appropriate coherence condition for compact closed categories,
%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
%% operational equivalence:

%% =======
\paragraph*{Coherence}
The coherence condition for compact closed categories may be visualized as
follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
%% >>>>>>> other
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
are to reverse the flow of execution, then this coherence condition is
satisfied.

@@

%% <<<<<<< local
%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
%% operations and this implies that {{eta+}}/{{eps+}} allow the
%% construction of a {{trace+}} operator (shown in Section
%% \ref{sec:monoidal-constructions}).  As shown in previous work
%% \cite{infeffects}, non-terminating computation can be constructed from
%%      {{trace+}} and recursive types.
%% =======
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

@@

\end{multicols}
%% >>>>>>> other

@@

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/neg_lift.pdf}
\end{center}

%subcode{proof} include main
%@ c : b1 <-> b2
%@@ neg c : (-b1) <-> (-b2)
\end{multicols}

@@

%% <<<<<<< local
%% %% The section follows the outline of the previous one, first presenting
%% %% the new syntax, and discussing the graphical language and the
%% %% properites implied by the underlying categorical structure.

%% =======
%% >>>>>>> other
Multiplicative duality is added to {{langRev}} by adding types
{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.

@@

For convenience we sometime use the notation {{b1/b2}} to indicate the
pair {{b1 * (1/b2)}}.  As with additive duals, we represent {{eta*}}
and {{eps*}} using U-shaped connectors. In later diagrams, the wires
for {{1}} will be dropped as they can freely introduced or eliminated
using {{uniti}} and {{unite}}.

@@

%% <<<<<<< local
%% The usual interpretation of {{b1*b2}} we have both a value of type
%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
%% to be viewed as a fission point for a value of type {{b}} and its
%% multiplicative inverse {{1/b}}. This naturally raises the question --
%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
%% not produce any specific value of type {{b}}, instead it produces a
%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
%% by the placeholder that indicates the absence of the unspecified
%% value.
%% =======
In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
and a value of type {{b2}}. This interpretation is maintained in the presence
of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
fission point for a value of type {{b}} and its multiplicative inverse
{{1/b}}. This naturally raises the question -- which value of type {{b}} is
produced? The isomorphism {{eta*}} does not produce any specific value of
type {{b}}, instead it produces a place holder for a value such that the
value of type {{1/b}} is always the dual of the chosen one.
%% >>>>>>> other

@@

%% <<<<<<< local
%% =======
%% >>>>>>> other

%% <<<<<<< local
In terms of \emph{logic programming}, {{eta* : 1 <-> (1/b)*b}} acts as
the site of creation of a new typed logic variable and constructs two
values using it: a value of type {{b}} and its dual of type
{{1/b}}. These two values are free to move independently through the
computation. Operations on one value affect what operations are possible
on the other since they are both related by an underlying logic
variable. The operator {{eps*}} acts as a unification site for values of
type {{b}} and {{1/b}} flowing along both branches. If the values are
not unifiable duals, then the specific program sequence fails. Like
with logic programming, one can has many worlds where some might fail,
while others make progress. Sec. \ref{sec:rat} has more details.  If
the values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

% \paragraph*{Entanglement and Anti-Particles.}

%% An analogy inspired by quantum mechanics is also applicable here.  The
%% operator {{eta*}} is a site of fission that creates \emph{an entangled
%%   particle and anti-particle}. These particles can be thought to be in
%% a \emph{superposition} of states determined by their type. Both flow
%% in the same direction in time.  In each possible world that the
%% particle exists, it takes on a specific value inhibiting its type and
%% correspondingly its entangled anti-particle takes the dual of the
%% specific value. Since the particles are entangled, actions on one
%% (such as transformation by application of an isomorphism) affects the
%% other in much the same sense as \emph{action at a distance}. The
%% `{{a-o*b}}' functions mentioned in the introduction, are first-class
%% values and they correspond to these entagled pairs. Finally {{eps*}}
%% operations are sites where corresponding particle and anti-particles
%% annihilate each other. While appealing as an analogy, this description
%% does not imply (or preclude) any formal connection with Physics.

%% We can ask the function to be applied to a value by
%% unifying the anti-particle with a particle in a well-known state and
%% examining is pair. We shall use this point of view in the development
%% our SAT solver (see Sec. \ref{sec:sat-solver}).

%%  This also leads to the exciting point of view that
%% entangled such particle/anti-particle pairs, which are first class
%% values, act as functions in a physical first-class sense.

%% Further the fission site does not create a particle in any specific
%% state, but in a superposition of all possible states as dictated by
%% its type.

% Since such pairs are first-class entities, they correspond to

%% \emph{Coherence for Compact Closure with Multiplicative Duality.}  As
%% before, the triangle axiom which is the coherence condition on compact
%% closed categories is applicable and is expressed below in terms of
%% {{eta*}}/{{eps*}} and the multiplicative monoid.
%% =======
%% >>>>>>> other

@@

\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}

\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

%% <<<<<<< local
Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.
%% =======
% \paragraph*{Unification.}
% A computational interpretation of this is that {{langRevEE}} acts like
% a logic programming system and {{eta*}} acts as the site of creation
% of a new typed logic variable of type {{b}} and its corresponding dual
% of type {{1/b}}. These two values are free to move independently
% through a circuit, but any operations on value affect what operations
% are possible on the other since they are both related by an underlying
% logic variable. Dually, {{eps*}} acts as a unification site for values
% flowing along both branches. If the values are not exactly duals, then
% the system fails to unify and hence the program is stuck. If the
% values are exactly duals, then unification succeeds and {{eta*}}
% returns {{1}}.

@@

% In a conventional language, higher-order functions capture scope and can be
% represented as first-class values. In {{langRevEE}}, we have two notions of
% functions both of which share similarities with conventional higher-order
% functions.
%% >>>>>>> other

@@

%% <<<<<<< local
\emph{Constructions.} The constructions for functions, involution, and trace are identical,
replacing {{+}} by {{*}}, \textit{mutatis mutandis}.

%% \begin{enumerate}
%% \item 
%% \emph{Trace}
%% =======
%% % \item
%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
%% % are indeed first-class values. However these functions can only be
%% % linearly used, i.e., they represent a constraint that can be satisfied
%% % at most once.
%% % %%  in one `computational world'. It may be possible that to
%% % %% repeatedly invoke multiplicative functions with an appropriate
%% % %% recursive type whose unfoldings allow for repeated unification.
%% % \end{itemize}
%% >>>>>>> other

@@

%% \item 
%% \emph{ {{eps_{fst} }} }

%% {{(

@@

To build these isomorphisms, we first build an intermeidate
construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/eps_fst.pdf}
}
\end{center}

The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
terms of {{eps_{fst} }} as shown below.

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/mult_neg.pdf}
}
\end{center}

The second isomorphism can be built in the same way by merely swapping
the arguments.

\item
\emph{Multiplying Negatives}

@@

This isomorphism can be built using \emph{involution} constructed in
Sec \ref{sec:neg} and the isomorphisms above, and correspond to the
algebraic manipulation:

%subcode{opsem} include main
%! columnStyle = rc
%   & b1 * b2
% = & -(-(b1*b2)) 
% = & -((-b1)*b2)
% = & (-b1)*(-b2)

@@

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT solvers
\amr{must cite ???} in that there is no explicit search operation on the
boolean space. It also varies from previous quantum SAT solvers \amr{must
  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
which is the solution following which an isomorphic clone operation allows us
to examine the result.  Though we focus on SAT here, this construction can be
generalized to any constraint satisfaction problem whose search space can be
represented by a recursive type and for which an isomorphic clone operation
can be constructed. Before we start, let us recall a constructions from 
prior work \cite{infeffects}.

\emph{Conditionals.}
Given any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
the combinator {{c}} is used to transform the value of type~{{b}}. If the
boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
write down the combinator for {{if_c}} in terms of {{c}} as 
{{ dist (;) ((id (*) c) (+) id) (;) factor }}.

% Let us look at the combinator pictorially as if it were a circuit and values
% are like particles that flow through this circuit.

The diagram below shows the input value of type {{(1+1)*b}} processed
by the distribute operator {{dist}}, which converts it into a value of
type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
case when the boolean is {{true}} (i.e. the value was {{left ()}}),
the combinator~{{c}} is applied to the value of type~{{b}}. The right
branch which corresponds to the boolean being {{false}} passes along
the value of type {{b}} unchanged.

\begin{center}
\scalebox{1.0}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
\includegraphics{diagrams/thesis/cnot.pdf}
}
\end{center}

% We will be seeing many more such wiring diagrams in this paper and it is
% useful to note some conventions about them. Wires indicate a
% value that can exist in the program. Each wire, whenever possible, is
% annotated with its type and sometimes additional information to help clarify
% its role. When multiple wires run in parallel, it means that those values
% exist in the system at the same time, indicating pair types. When there is a
% disjunction, we put a {{+}} between the wires. 
% Combinators for distribution {{dist}} and factoring {{factor}}
% are represented as triangles with their operator symbols in them. Other
% triangles may be used and, in each case, types or labels will be used to
% clarify their roles. Finally, we don't draw boxes for combinators such as
% {{id}}, commutativity, and associativity, but instead just shuffle the wires
% as appropriate.

\emph{Boolean Operation based on {{if_c}}.}
The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only is the first is {{false}}.

Similarly, we can iterate the construction of {{if_c}} to check
several bits. The gate {{if_{cnot} }}, which we may also write as
{{if^2_{not} }}, checks two booleans and negates the result wire only
if they are both {{true}}. The gate {{if^2_{not} }} is well known as
the universal reversible primitive gate called the `Toffoli gate'
\cite{Toffoli:1980}.  We can generalize this construction to
     {{if^n_{not} }} which checks {{n}} bits and negates the result
     wire only if they are all {{true}}.

\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
property that as long as the second argument is fixed to be {{true}},
output of the second wire is always the same as teh value of the first
wire. In other words {{else_{not} }} can be thought of as cloning the
value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
inputs in the process. Let us call this construction 
{{clone^n_{bool} }}.

%% \begin{enumerate}
%% \item Recall from previous work that for the types {{bool}} and, by
%%   extension, any tuple of {{bool}}, an isomorphic clone operation can
%%   be constructed (See Lemma 7.2 \cite{infeffects}).

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
%%   {{if_{not} }}, which when generalized to two bits was
%%   {{toffoli}}. This construction maybe generalised to {{n}} bits as
%%   {{if^n_{not} }}.
%% \end{enumerate}

@@

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match, i.e., the dotted part in the
diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
the control wire.

@@

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.

@@

\paragraph*{QC} 
A \emph{time traveling} intuition is also applicable here.  In a
normal cicruit, as computational steps are taken values flow from left
to right, i.e. computational time progresses from left to right.  The
action of {{eps+}} causes values, aka information particles, to flow
from right to left i.e. backwards in computational time.  The
interesting thing about this interpretation is that particles that
travel backwards in time get to see and interact with the history of
particles that they coexist with. This gives us an intuitive
interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
Sec. \ref{sec:specific-constructions}) where it would otherwise seem
that {{-b1}} and {{b2}} move in opposite directions. The backward flow
of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
equivalent to both particles moving backward in time.

An analogy inspired by quantum mechanics is also applicable here.  The
operator {{eta*}} is a site of fission that creates \emph{an entangled
  particle and anti-particle}. These particles can be thought to be in
a \emph{superposition} of states determined by their type. Both flow
in the same direction in time.  In each possible world that the
particle exists, it takes on a specific value inhibiting its type and
correspondingly its entangled anti-particle takes the dual of the
specific value. Since the particles are entangled, actions on one
(such as transformation by application of an isomorphism) affects the
other in much the same sense as \emph{action at a distance}. The
`{{a-o*b}}' functions mentioned in the introduction, are first-class
values and they correspond to these entagled pairs. Finally {{eps*}}
operations are sites where corresponding particle and anti-particles
annihilate each other. While appealing as an analogy, this description
does not imply (or preclude) any formal connection with Physics.

Are constructions are very similar to those developed in the context
of QC. Is this how nature computes? Are the analogies to time
traveling particles and anti-particles more concrete in some way?

>>>>

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}}, say \$20.00, and eventually
produces another \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted line
with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type {{a+0}}; we indicate this value
  as {{left 20}};
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
%% The example illustrates that the empty type and negative types have a
%% computational interpretation related to continuations: negative types denote
%% values that backtrack to satisfy dependencies, or in other words act as debts
%% that are satisfied by the backward flow of information.

@@

As before, the execution begins at line~(0) with the value {{20}}. At
line~(1) two values, {{20}} and {{()}}, flow forward. One can think of the
value {{()}} (of type {{1}}) as ``having a credit card.'' The credit card
isn't money, nor is it debt, but is the option to generate a credit-debt
constraint.  At line~(2) we exercise this option and hence have three values:
the initial value {{20}} flowing from line (1) and two entangled values,
{{1/alpha}} and {{alpha}}. The {{alpha}} and {{1/alpha}} are unspecified
values, i.e., we don't yet know how much money we need to borrow, but we do
know that what is borrowed must be what is returned. Hence~{{alpha}} denotes
the presence of an unknown quantity and dually {{1/alpha}} should be thought
of as the absence of an unknown quantity.
% Unlike with negative types, wherein only one value existed at a time and the
% computation backtracked, here we have three values that \emph{exist at the
%   same time}. In other words, the computation with fractions is realized with
% a schedule in which every value independently and concurrently proceeds
% through its subcomputation.
At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
fixing values of all three branches, i.e., we have paid our debt of the missing
information and the third subcomputation has the known value {{20}}. After
payment of our debt, as line (4) shows, we have use of our credit card again.

@@

\noindent 
\textit{Example.} The following combinator is represented by the given
diagram:

{{c : b * (1+1) <-> b + b}}

@@

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

As will be detailed in the remainder of this section, the compact closed
structure provides several properties of interest: (i) morphisms or wires are
allowed to run from right to left, (ii) the structure admits a trace that,
depending on the situation, can be used to implement various notions of loops
and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual operator is
an involution, and (iv) the structure is equipped with objects representing
higher-order functions and a morphism \textit{eval} that applies these
functional objects. Interestingly each monoidal structure provides the same
ingredients, resulting in two dualities, two traces, two involutions, and two
notions of higher-order functions.

@@

\paragraph*{Negative Information Flow.} 
The intuitive way to read the diagrams is that they reverse the flow of
information. In previous {{langRev}} circuits, information was always flowing
from left to right. The presence of {{eps+}} operations cause information to
flow from right to left which can be reversed again using {{eta+}}.  As an
example, the diagram representing the isomorphism from the introduction is:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/algebra1.pdf}
}
\end{center}

% \paragraph*{Time Traveling.}
% A time traveling intuition is also applicable here: if view the
% left-to-right direction of information flow as the canonical forward
% direction of information, then the action of {{eps+}} allows values,
% aka information particles, to flow backwards in time. Particles that
% flow backwards in time get to see and interact with the history of
% particles that they coexist with, i.e., values that they exist
% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
% (whose construction is shown in Section \ref{sec:neg-constructions}).

%\subsection{Compact Closure with Additive Inverses}
\paragraph*{Coherence}
The coherence condition for compact closed categories may be visualized as
follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
are to reverse the flow of execution, then this coherence condition is
satisfied.

@@

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

@@

\end{multicols}

@@

Given {{c : b1 <-> b2}}

\begin{center}
  \includegraphics{diagrams/neg_lift.pdf}
\end{center}

@@

Multiplicative duality is added to {{langRev}} by adding types
{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.

@@

As with additive duals, we represent {{eta*}} and {{eps*}} using
U-shaped connectors. In later diagrams, the wires for {{1}} will be
dropped as they can freely introduced or eliminated using {{uniti}}
and {{unite}}.

@@

In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
and a value of type {{b2}}. This interpretation is maintained in the presence
of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
fission point for a value of type {{b}} and its multiplicative inverse
{{1/b}}. This naturally raises the question -- which value of type {{b}} is
produced? The isomorphism {{eta*}} does not produce any specific value of
type {{b}}, instead it produces a place holder for a value such that the
value of type {{1/b}} is always the dual of the chosen one.

@@

@@

% \paragraph*{Unification.}
% A computational interpretation of this is that {{langRevEE}} acts like
% a logic programming system and {{eta*}} acts as the site of creation
% of a new typed logic variable of type {{b}} and its corresponding dual
% of type {{1/b}}. These two values are free to move independently
% through a circuit, but any operations on value affect what operations
% are possible on the other since they are both related by an underlying
% logic variable. Dually, {{eps*}} acts as a unification site for values
% flowing along both branches. If the values are not exactly duals, then
% the system fails to unify and hence the program is stuck. If the
% values are exactly duals, then unification succeeds and {{eta*}}
% returns {{1}}.

@@

% In a conventional language, higher-order functions capture scope and can be
% represented as first-class values. In {{langRevEE}}, we have two notions of
% functions both of which share similarities with conventional higher-order
% functions.

@@

% \item
% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
% are indeed first-class values. However these functions can only be
% linearly used, i.e., they represent a constraint that can be satisfied
% at most once.
% %%  in one `computational world'. It may be possible that to
% %% repeatedly invoke multiplicative functions with an appropriate
% %% recursive type whose unfoldings allow for repeated unification.
% \end{itemize}

@@

\item 
\emph{ {{eps_{fst} }} }

{{(-b1)*b2 + b1*b2 <-> 0}}

\begin{center}
  \includegraphics{diagrams/eps_fst.pdf}
\end{center}

@@

\begin{center}
  \includegraphics{diagrams/mult_neg.pdf}
\end{center}

The following isomorphism can be constructed similarly

@@

@@

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT solvers
\amr{must cite ???} in that there is no explicit search operation on the
boolean space. It also varies from previous quantum SAT solvers \amr{must
  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
which is the solution following which an isomorphic clone operation allows us
to examine the result.  Though we focus on SAT here, this construction can be
generalized to any constraint satisfaction problem whose search space can be
represented by a recursive type and for which an isomorphic clone operation
can be constructed. Before we start, let us recall two constructions:

\begin{enumerate}
\item Recall from previous work that for the types {{bool}} and, by
  extension, any tuple of {{bool}}, an isomorphic clone operation can
  be constructed (See Lemma 7.2 \cite{infeffects}).

\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
  {{if_{not} }}, which when generalized to two bits was
  {{toffoli}}. This construction maybe generalised to {{n}} bits as
  {{if^n_{not} }}.
\end{enumerate}

@@

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match, i.e., the dotted part in
the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
`heap control wire'.

@@

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs'.

@@

\paragraph*{QC} 
Are constructions are very similar to those developed in the context
of QC. Is this how nature computes? Are the analogies to time
traveling particles and anti-particles more concrete in some way?
